DNNNNR0: How to Avoid OOTA Without Really Trying

### 2025-03-27 paper/ootavacuous.tex
ootavacuous: Add citations for litmus tests
The unspecified ordering for ++ might be *somewhere*...

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Simple reordering can cause non-OOTA, but OOTA-like, behavior, as
exemplified by
Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
Either the compiler or the CPU might reorder T1's lines~1 and~2, so that
all of \co{x}, \co{y}, \co{r1}, and \co{r2} might end up with the value 42,
as indicated by the ``sometimes satisfied'' in the figure.

@@

On the other hand, OOTA cycles need not involve nonsensical or
unmotivated values, as shown in
Listing~\ref{lst:OOTA Cycle}.
Line~3 of both threads stores the value 42 only when line~2 determines
that the value loaded was 42.
As with Listing~\ref{lst:Simple OOTA}, hardware ordering prevents the
OOTA result, as discussed further in
Section~\ref{sec:Hardware Dependencies, Instruction Ordering, and the Fundamental Property}.

@@

All OOTA researchers owe a debt of gratitude to the foundational work led
by William Pugh, represented by the infamous Java-language
``Causality Test Cases''\footnote{
	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}~\cite{JeremyManson2005JavaMemoryModel}.
The OOTA problem has been open for more than 20~years and has been shown
to be quite difficult~\cite{MarkBatty2015OOTAisHard}.
% %%% @@@ Eventually cite P3064.

@@

Semantic dependencies form only one type of link in an OOTA cycle.
The other type extends from a given store to a load that returns the
value stored.
It is tempting to argue that the store must precede
the load in global time and combine this with the intuitive notion that
any real C++ implementation must consume global time when computing a
semantic dependency.
This combination suggests that OOTA cycles cannot occur.
The idea has been formalized by defining an OOTA cycle as a cycle
in sdep $\cup$ rf~\cite{PaulEMcKenney2014OOTA},
% This citation credits the formulation to Ali Sezgin, third author
% of N4323 ("Out-of-Thin-Air Execution is Vacuous").
where sdep is the set of semantic dependencies within each thread and
rf is the set of store-to-load ``reads-from'' links, whether internally
within a thread (rfi) or externally between threads (rfe).
% @@@ Consider adding citation defining rf, rfi, and rfe.

@@

Compare this example to the previous one:
\begin{quote}
\begin{verbatim}
if (x > 0)
    y = 42;
else
    y = 42;
\end{verbatim}
\end{quote}
Because the stores executed on each arm of the \co{if} statement write
identical values to identical addresses, one could equally well regard
the two statements as performing two different stores or as performing
for all intents and purposes a single store, independent of \co{x}.
Reasonable C++ implementations might disagree on this matter and
therefore on whether or not the example has a semantic dependency.
It is the implementation's choice.

@@

Consider the following example, with variable \co{i} initially zero:
\begin{quote}
\begin{verbatim}
int foo(int a, int b)
{
   return a / b;
}

@@

Invented atomic loads are problematic.
Consider this code:
\begin{quote}
\begin{verbatim}
int r0 = x;
int r1 = r0 * r0 + 2 * r0 + 1;
\end{verbatim}
\end{quote}
For values of \co{x} small enough to avoid overflow, the abstract machine is
guaranteed to produce a perfect square in \co{r1}.

@@

Inventing atomic loads can also destroy dependencies.
For example, consider this code:
\begin{quote}
\begin{verbatim}
   int r1 = (x != 0);
   int r2 = (y != 0);
   z = (r1 == r2);
\end{verbatim}
\end{quote}
It seems clear that the store to \co{z} semantically depends on the load
from \co{y}, because the value of \co{z} will change whenever \co{y}
changes between zero and nonzero (all else being equal).

>>>>

Simple reordering can cause non-OOTA, but OOTA-like, behavior, as
exemplified by
Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2014OOTA}.
Either the compiler or the CPU might reorder T1's lines~1 and~2, so that
all of \co{x}, \co{y}, \co{r1}, and \co{r2} might end up with the value 42,
as indicated by the ``sometimes satisfied'' in the figure.

@@

On the other hand, OOTA cycles need not involve nonsensical or
unmotivated values, as shown in
Listing~\ref{lst:OOTA Cycle}, adapted from~\cite{PaulEMcKenney2014OOTA}.
Line~3 of both threads stores the value 42 only when line~2 determines
that the value loaded was 42.
As with Listing~\ref{lst:Simple OOTA}, hardware ordering prevents the
OOTA result, as discussed further in
Section~\ref{sec:Hardware Dependencies, Instruction Ordering, and the Fundamental Property}.

@@

All OOTA researchers owe a debt of gratitude to the foundational work led
by William Pugh, represented by the infamous Java-language
``Causality Test Cases''\footnote{
	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}~\cite{JeremyManson2005JavaMemoryModel}.
The OOTA problem has been open for more than 20~years and has been shown
to be quite difficult~\cite{MarkBatty2015OOTAisHard,PaulEMcKenney2023P3046R2/OOTAwithoutTrying}.

@@

Semantic dependencies form only one type of link in an OOTA cycle.
The other type extends from a given store to a load that returns the
value stored.
It is tempting to argue that the store must precede
the load in global time and combine this with the intuitive notion that
any real C++ implementation must consume global time when computing a
semantic dependency.
This combination suggests that OOTA cycles cannot occur.
The idea has been formalized by defining an OOTA cycle as a cycle
in sdep $\cup$ rf~\cite{PaulEMcKenney2014OOTA},
% This citation credits the formulation to Ali Sezgin, third author
% of N4323 ("Out-of-Thin-Air Execution is Vacuous").
where sdep is the set of semantic dependencies within each thread and
rf is the set of store-to-load ``reads-from'' links, whether internally
within a thread (rfi) or externally between threads (rfe).\footnote{
	Additional background on temporal distinctions between coe, fre,
	and rfe maybe found elsewhere~\cite[Appendix A]{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}.}

@@

Compare this example~\cite{Howells2009membartxt} to the previous one:
\begin{quote}
\begin{verbatim}
if (x > 0)
    y = 42;
else
    y = 42;
\end{verbatim}
\end{quote}
Because the stores executed on each arm of the \co{if} statement write
identical values to identical addresses, one could equally well regard
the two statements as performing two different stores or as performing
for all intents and purposes a single store, independent of \co{x}.
Reasonable C++ implementations might disagree on this matter and
therefore on whether or not the example has a semantic dependency.
It is the implementation's choice.

@@

Consider the following example, with variable \co{i} initially zero:
% @@@ Citation?
\begin{quote}
\begin{verbatim}
int foo(int a, int b)
{
   return a / b;
}

@@

Invented atomic loads are problematic.
Consider this code~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}:
\begin{quote}
\begin{verbatim}
int r0 = x;
int r1 = r0 * r0 + 2 * r0 + 1;
\end{verbatim}
\end{quote}
For values of \co{x} small enough to avoid overflow, the abstract machine is
guaranteed to produce a perfect square in \co{r1}.

@@

Inventing atomic loads can also destroy dependencies.
For example, consider this code~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}:
\begin{quote}
\begin{verbatim}
   int r1 = (x != 0);
   int r2 = (y != 0);
   z = (r1 == r2);
\end{verbatim}
\end{quote}
It seems clear that the store to \co{z} semantically depends on the load
from \co{y}, because the value of \co{z} will change whenever \co{y}
changes between zero and nonzero (all else being equal).
